---
title: "Lab2 - inferential statistics - confidence intervals"
author: "Isabel Sassoon"
date: "07/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Confidence Intervals in R

We will be showing you how to use R to compute the confidence intervals for a proportion and for a mean.

## Confidence interval for a proportion

A Market research agency interviews a random sample of 1000 people who live in London, and finds that 49% of people suggest they will vote Labour for the London Mayor. Find a 95% and a 90% confidence interval for the true proportion of voters who will vote for a Labour Mayor.

```{r}
prop.test(490, 1000, conf.level = 0.95)
```

95 percent confidence interval:
 0.46 0.52


## What about the 90% CI?


```{r}
prop.test(490, 1000, conf.level = 0.9)
```

The 90% confidence interval is

 0.46 0.52
 
In this case when apprximated to 2 decimal places the confidence intervals are almost identical! 

## Confidence interval for the mean

### Reading in from csv - use the skew data.csv

```{r}
data<-read.csv("data/skewdata.csv")
```

## Visualise the data

```{r}
summary(data)
```

This data contains only one variable. We can see its mean is 30.969.


# What is the 95% confidence interval for this variable?

```{r}
t.test(data$values)
```

The 95 percent confidence interval: is 24.11 37.82
 
## We can do the same for the 90 percent confidence interval

```{r}
t.test(data$values, conf.level = 0.9)
```
90 percent confidence interval: 25.27 36.66

We can see that the confidence interval is smaller as we decrease our confidence level. 

***

**The section below has the calculation broken down into the equation if you are interested in the mechanics of the computation**

## How to run confidence intervals for this variable - Z and T at the 95% confidence level

```{r}
mean.values<-mean(data$values)
sd.values<-sd(data$values)
```

In order to compute the confidence interval for a mean using Z we would apply this equation

: $$  (\bar{x} -  Z \times \frac{s}{\sqrt{n}}, \bar{x} + Z \times \frac{s}{\sqrt{n}}) $$

We have the value for $\bar{x}$ and s from the code chunk above.
We can get the value of Z 0.95 from R 

```{r}
qnorm(0.975)
```

Recall 0.95 confidence for a two sided test will result in us looking for Z 0.975 on each side. The number is the same and it is 1.96

Then we just need to combine this to compute the intervals

```{r}
ucl<-mean.values+qnorm(0.975)*sd.values/sqrt(30)
lcl<-mean.values-qnorm(0.975)*sd.values/sqrt(30)

```

```{r}
ucl
lcl
```

# If we replace Z with T

```{r}
qt(0.975, 29)
```

```{r}
uclt<-mean.values+qt(0.975,30)*sd.values/sqrt(30)
lclt<-mean.values-qt(0.975,30)*sd.values/sqrt(30)
```

```{r}
uclt
lclt
```

The confidence levels are very similar to the ones obtained with Z.

