---
title: "Lab2 - inferential statistics - Probability distributions"
author: "Isabel Sassoon"
date: "07/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability functions

There are four basic probability functions for each probability distribution in R. These begin with the same four prefixes: d, p, q or r followed by the root name of the probability that identifies the specific distribution. For example for the **normal** distribution the name to use is **norm**.

 - d is for *”density”* and the corresponding function returns the value from the probability density function (continuous) or probability mass function (discrete).
 - p is for *”probability”* and the corresponding function returns a value from the cumulative distribution function.
 - q is for *”quantile”* and the corresponding function returns a value from the inverse cumulative distribution function.
 - r is for *”random"* and the corresponding function returns a value drawn randomly from the given distribution

## The normal distribution
 
When we need to work with different functions related to the normal distribution then the following syntax is used

To obtain the **density** of the normal distribution:

d+ name of the distribution (the value on the x axis)

So for example if we wanted to know the value of y (from the y axis), from the normal distribution curve that corresponds to x=2 then we would use:

```{r}
dnorm(2)
```


 If we wanted the value of y, but using a normal distribution with different means and s.d these are optional arguments in the function:
 
```{r}
dnorm(11,10,0.5)
```

*pnorm* is the cumulative distribution function for the normal distribution. By definition $pnorm(x) = P(X \leq x)$ and is the area under the normal density curve to the left of x. 

```{r}
pnorm(2)
```


The value above shows pnorm(2), the area under the normal density curve to the left of $x = 2$. From this we can see that the probability that a standard normal random variable takes on a value less than or equal to 2 is 0.977

*qnorm* is the quantile function of the standard normal distribution. If $qnorm(x)= k$ then k is the value such that $P(X \leq k) = x$. *qnorm* is the inverse function for *pnorm*. 

```{r}
qnorm(0.977)
```


From the chunk of code above  we have: qnorm(0.977) = qnorm(pnorm(2)) = 2.

*rnorm* generates random values from a standard normal distribution. The required argument is a number specifying the number of normal variables to produce. 

```{r}
x<-rnorm(20)
x1<-rnorm(2000)
```


The figure below illustrates rnorm(20) and rnorm(2000) the larger the number in the bracket the more "bell curved" the plot is.

```{r}

hist(x)
hist(x1)
```

***

# Examples using other distributions

## Poisson Distribution

```{r}

x1=0:100
y2<-dpois(x1,lambda=50)

plot(x1,y2, type="h", main = "Poisson Distribution (Lambda=50)", ylab="Frequency", xlab="Count")
```


## Binomial Distribution

```{r}
y<-dbinom(x, size=10, prob = 0.5)

plot(x,y,type='h',
     main='Binomial Distribution (n=10, p=0.5)',
     ylab='Probability',
     xlab ='# heads',
     lwd=3)
```

## Uniform Distribution

```{r}
xu<-runif(100000, min=1, max=2)
xu<-as.data.frame(xu)
ggplot(xu, aes(x=xu)) + geom_density() + theme_classic() +
  labs(main="A Uniform Distribution", xlab="value", ylab="Density") + xlim(0,3)
```


## Exponential Distribution


```{r}
#exponential distribution

xe<-rexp(10000, rate = 10)
xe<-as.data.frame(xe)

ggplot(xe, aes(x=xe)) + geom_density() + theme_classic() +
  labs(main="An Exponential Distribution", xlab="value", ylab="Density") + xlim(0,3)
```